\documentclass[12pt]{article}
\usepackage{fontspec}
\setmainfont{EB Garamond}[Numbers=OldStyle, Ligatures=TeX]
\usepackage[letterpaper, margin=1.25in]{geometry}
\usepackage{hyperref}

\begin{document}
\thispagestyle{empty}

\noindent\today

\bigskip

\noindent Dear Editors,

\medskip

I am submitting ``What Dogs Know About Trails: Metaphor, Mechanisms, and the Phylogeny of Cross-Domain Extension'' for consideration as a research article in \textit{Metaphor and Symbol}.

The paper argues that Boyd's Homeostatic Property Cluster (HPC) theory and Lakoff and Johnson's Conceptual Metaphor Theory (CMT) are natural complements that have developed without mutual engagement. The synthesis reframes metaphor as cross-domain HPC extension on partial property overlap, providing what each framework lacks: HPC supplies the stability mechanisms CMT needs; CMT reveals a cross-domain capacity HPC hasn't theorized. The reframing recasts standing disputes (comparison vs.\ categorization, embodied vs.\ abstract, attribute vs.\ relational mapping) and generates three testable predictions, including that any animal with HPC-structured categories should show cross-domain recruitment.

I believe \textit{Metaphor and Symbol} is the right venue for this work because it bridges metaphor theory with the philosophy of science tradition that metaphor researchers have largely overlooked. Boyd's theory-constitutive metaphors and his HPC framework address questions central to the journal's readership (what makes metaphors productive, why some conventionalize, what distinguishes literary from everyday from constitutive metaphor), but from a direction that hasn't been integrated with Lakoff and Johnson's programme. The paper also extends the framework to non-human cognition and LLM behaviour, generating testable predictions at each level.

This is a theoretical paper with no empirical data. The manuscript is approximately 9,500 words (excluding references). It has not been submitted elsewhere. A preprint (v1) is available on PsyArXiv at \url{https://osf.io/preprints/psyarxiv/vuafy_v1}. As sole author, I confirm that I have approved the final submission.

I am happy to provide reviewer suggestions if helpful.

\bigskip

\noindent Sincerely,\\
Brett Reynolds\\
Humber Polytechnic \& University of Toronto\\
\href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}

\end{document}
